A model-based method for measuring stability in personal culture with panel data[^thanks]

### 2020-12-18 ms.Rmd
updated figure, minor additions

@@

While we cannot account for attitudes and behaviors that are not measured here, these results suggest, consistent with KV, that a wide range of attitudes and behaviors tend to be settled by the time people are old enough to participate in surveys like the GSS. While this is not surprising since we examine the same set of questions as KV, it is important that different statistical assumptions lead to similar substantive conclusions. People might change their responses from wave to wave, but this change does not tend to persist, meaning it reflects temporary changes in opinion or measurement error.[^temp] There is little evidence of even a moderate amount of durable change in opinions in the medium-term (2-4 years). Perhaps even more striking, the results in Section 4.2 show that the majority of variables are consistent with no change at all over the four-year period.

@@

On specific variables, the approach presented here occasionally suggests different conclusions than those presented in KV, as would be expected under models that make different assumptions, with the model presented here favoring the SDM model in more cases. There are two main reasons for these differences. First, KV pool their panels together, meaning that if _any_ of the three panels show updating, the variable as a whole is likely to show updating. A second these differences emerge is because the model presented here uses slightly more data than the KV model. A person who gives the same response in waves 1 and 2 but a different response in wave 3 is mostly useless under the KV model but can factor into the calculation in the model presented here. Because variable-panels can switch from favoring the active updating model to favoring the settled dispositions model with the addition or subtraction of a few cases, this suggests that evidence of the AUM is very weak even when it is favored. In particular, when the AUM is favored, it is because a very small proportion of the population is making durable changes. It is difficult to draw broader conclusions about these discrepancies, as they cut across both question substance and question structure.

@@

Similarly, KV suggested that public behaviors were more likely to demonstrate active updating, since these behaviors receive social reinforcement that facilitate durable change. In the results presented here, social behaviors such as socializing at a bar and religious behaviors such as church attendance are the two categories that show the highest rates of active updating, with 41 percent and 50 percent of variable-panels in these categories, respectively, showing evidence of active updating.

@@

These findings have clear implications for cultural sociology and public opinion research. We believe that these findings are also relevant for research on cultural evolution [see @henrich2015; @heyes2018; @mesoudi2016] because they place significant limits on the likely pace and mechanisms of cultural change. Cohort-based change will necessarily be slower than that implied by simple population models of social learning. This lag process allows cultural equilibria to be durable even when the contemporary information environment is changing [@oconnor2019]. We hope future research explores these links more directly.

>>>>

An additional advantage of this approach is that can be more easily extended to panel data sets that include a larger number of waves than the model presented in KV. While three waves is the minimum number required to adjudicate the two theoretical approaches outlined above, and what we focus on in this paper, adding more waves can help adjudicate the two theoretical models under a wider array of assumptions.

@@

While we cannot account for attitudes and behaviors that are not measured here, these results suggest, consistent with KV, that a wide range of attitudes and behaviors tend to be settled by the time people are old enough to participate in surveys like the GSS. While this is not surprising since we examine the same set of questions as KV, it is important that different statistical assumptions lead to similar substantive conclusions.

It is important to note here that settled opinions does not necessarily mean stable (unchanging) opinions. People might change their responses from wave to wave, but the settled dispositions model assumes this change does not tend to persist, meaning it reflects temporary changes in opinion or measurement error.[^temp] There is little evidence of even a moderate amount of durable change in opinions in the medium-term (2-4 years). Perhaps even more striking, the results in Section 4.2 show that the majority of variables are consistent with no population-level change in attitudes or behaviors at all over the four-year period.

@@

On specific variables, the approach presented here occasionally suggests different conclusions than the approach presented in KV, as would be expected under models that make different assumptions, with the model presented here favoring the SDM model in more cases. There are two main reasons for these differences. First, KV pool their panels together, and the method favors the active updating model is any updating is present, meaning that if _any_ of the three panels show updating, the variable as a whole is likely to show updating. A second reason these differences emerge is because the model presented here uses slightly more data than the KV model. A person who gives the same response in waves 1 and 2 but a different response in wave 3 is mostly useless under the KV model but can factor into the calculation in the model presented here. Because variable-panels can switch from favoring the active updating model to favoring the settled dispositions model with the addition or subtraction of a few cases, this suggests that evidence of the AUM is very weak even when it is favored. In particular, when the AUM is favored, it is often because a very small proportion of the population is making durable changes. It is difficult to draw broader conclusions about these discrepancies, as they cut across both question substance and question structure.

@@

Similarly, KV suggested that public behaviors were more likely to demonstrate active updating, since these behaviors receive social reinforcement that facilitates durable change. In the results presented here, social behaviors such as socializing at a bar and religious behaviors such as church attendance are the two categories that show the highest rates of active updating, with 41 percent and 50 percent of variable-panels in these categories, respectively, showing evidence of active updating.

@@

These findings have clear implications for cultural sociology and public opinion research, favoring models of cultural change rooted in cohort replacement [@mannheim1952; @ryder1965] and those that place importance on early-life socialization over contemporary social environments [such as @bourdieu1990]. We believe that these findings are also relevant for research on cultural evolution [see @henrich2015; @heyes2018; @mesoudi2016] because they place significant limits on the likely pace and mechanisms of cultural change. Cohort-based change will necessarily be slower than that implied by simple population models of social learning. This lag process allows cultural equilibria to be durable even when the contemporary information environment is changing [@oconnor2019]. We hope future research explores these links more directly.
